On the surprising effectiveness of a simple matrix exponential derivative approximation, with application to global SARS-CoV-2

### 2023-09-14 manuscript/pnas_revision.tex
more revisions

We turn to our probabilistic bounds on the singular values of randomly
generated rate matrices, \cref{t:lambda_n-1(Q)->-infty_non-symm_a.s.}.
Although interesting in its own right, this result leads to
consequences for the bounds in \eqref{eq:qetq:exp:decay} and
\eqref{eq:ass:error} when applied in \cref{thm:bride:of:frankinstein}.
Before proceeding, we briefly introduce further mathematical
preliminaries associated with the so-called sub-exponential random
variables.  To avoid confusion, note that the following definition uses the term in the same way as, e.g., \cite{vershynin:2018}, but that other definitions that mean quite the opposite (i.e., heavier than exponential tails) appear in the literature \cite{goldie1998subexponential}.
\begin{definition}
  \label{def:sub:exp:RV}
  A random variable $X$ is called \textit{sub-exponential} if there
  exists some constant $K > 0$ for which its tails satisfy
\begin{equation}\label{e:def_subexp_RV}
\bbP(|X|\geq t) \leq 2 \exp\{-t/K\}, \quad \forall t \geq 0.
\end{equation}
In this case, the \textit{sub-exponential norm} of $X$ is defined by
\begin{equation}\label{e:sub-exp_norm}
\|X\|_{\psi_1} = \inf\big\{s > 0: \bbE \exp(|X|/s) \leq 2\big\}.
\end{equation}
The \textit{class of sub-exponential distributions} is denoted by
\begin{align*}
L_{\psi_1} = \big\{F_X(dx): \|X\|_{\psi_1} < \infty \big\}.
\end{align*}
\end{definition}

\begin{remark}\label{rmk:exp:mom}
  In fact, by \cite[Proposition 2.7.1, p.\ 33]{vershynin:2018},
  condition \eqref{e:def_subexp_RV} is equivalent to the existence of
  some $s_0 > 0$ such that $\bbE \exp(|X|/s_0) \leq 2$, namely,
  $\|X\|_{\psi_1} < \infty$ in \eqref{e:sub-exp_norm}.  As a notable
  example, if $X \sim \exp(\lambda)$, $\lambda > 0$, then it is easy
  to see that $X \in L_{\psi_1}$, indeed.
\end{remark}

@@

\begin{remark}\label{rmk:non:sym:num:ev}
  Our rigorous formulation of \cref{cor:rig:det:rand:wrapup} is
  limited to symmetric random rate matrices whose above diagonal
  elements are independent and identically distributed draws from a
  sub-exponential distribution. However, strong numerical evidence
  suggest that the scope of the approximations
  \eqref{eq:bride:of:frankinstein:1:rm:app},
  \eqref{eq:bride:of:frankinstein:0:rm:app} reach far beyond
  limitations of \cref{cor:rig:det:rand:wrapup} in several different
  ways. \cref{fig:app} explores the
  consequences of relaxing various assumptions of
  \cref{cor:rig:det:rand:wrapup}.  The third plot involves folded Cauchy random variables, the heavy tails of which violate the sub-exponential assumption (\cref{def:sub:exp:RV}).  In general, there appears to be no
  significant departure from the idea that the proxy form
  $t(\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+}$ corresponds increasingly well to
  the true approximation error (\eqref{eq:true:diff}) as the dimension
  increases.
\end{remark}

@@

In addition to these scientific questions of interest, we are
interested in the performance of the first-order approximation as a
surrogate gradient for HMC in such a high-dimensional setting. To
isolate this performance, we fix the Bayesian bridge global-scale
$\tau$ at 2.5$\times10^{-5}$.  We generate a Markov chain with 80,000
states, saving every tenth state and removing the first 1,000 states as
burn-in.  Despite the relatively small number of iterations, we
observe large ESS that suggest satisfactory accuracy of the
first-order approximation within the context of high-dimensional
surrogate HMC.  The ESS for the three fixed-effect regression
coefficients $\theta_1$, $\theta_2$ and $\theta_3$ are 1,053, 1,343
and 498, respectively.  The median and minimum ESS for the 1,892
random effects $\epsilon_{ij}$ are 1,514 and 1,161, respectively.

>>>>

We turn to our probabilistic bounds on the singular values of randomly
generated rate matrices, \cref{t:lambda_n-1(Q)->-infty_non-symm_a.s.}.
Although interesting in its own right, this result leads to
consequences for the bounds in \eqref{eq:qetq:exp:decay} and
\eqref{eq:ass:error} when applied in \cref{thm:bride:of:frankinstein}.
Before proceeding, we briefly introduce further mathematical
preliminaries associated with the so-called sub-exponential random
variables.  To avoid confusion, note that the following definition uses the term in the same way as, e.g., \cite{vershynin:2018}, but that other definitions that mean quite the opposite (i.e., heavier than exponential tails) appear in the literature \cite{goldie1998subexponential}.
\begin{definition}
  \label{def:sub:exp:RV}
  A random variable $X$ is called \textit{sub-exponential} if there
  exists some constant $K > 0$ for which its tails satisfy
\begin{equation}\label{e:def_subexp_RV}
\bbP(|X|\geq t) \leq 2 e^{-t/K}, \quad \forall t \geq 0.
\end{equation}
In this case, the \textit{sub-exponential norm} of $X$ is defined by
\begin{equation}\label{e:sub-exp_norm}
\|X\|_{\psi_1} = \inf\big\{s > 0: \bbE e^{|X|/s} \leq 2\big\}.
\end{equation}
The \textit{class of sub-exponential distributions} is denoted by
\begin{align*}
L_{\psi_1} = \big\{F_X(dx): \|X\|_{\psi_1} < \infty \big\}.
\end{align*}
\end{definition}

\begin{remark}\label{rmk:exp:mom}
  In fact, by \cite[Proposition 2.7.1, p.\ 33]{vershynin:2018},
  condition \eqref{e:def_subexp_RV} is equivalent to the existence of
  some $s_0 > 0$ such that $\bbE e^{|X|/s_0} \leq 2$, namely,
  $\|X\|_{\psi_1} < \infty$ in \eqref{e:sub-exp_norm}.  As a notable
  example, if $X \sim \exp(\lambda)$, $\lambda > 0$, then it is easy
  to see that $X \in L_{\psi_1}$, indeed.
\end{remark}

@@

\begin{remark}\label{rmk:non:sym:num:ev}
  Our rigorous formulation of \cref{cor:rig:det:rand:wrapup} is
  limited to symmetric random rate matrices whose above diagonal
  elements are independent and identically distributed draws from a
  sub-exponential distribution. However, strong numerical evidence
  suggest that the scope of the approximations
  \eqref{eq:bride:of:frankinstein:1:rm:app},
  \eqref{eq:bride:of:frankinstein:0:rm:app} reach far beyond
  limitations of \cref{cor:rig:det:rand:wrapup} in several different
  ways. \cref{fig:app} explores the
  consequences of relaxing various assumptions of
  \cref{cor:rig:det:rand:wrapup}.  The third plot involves folded Cauchy random variables, the heavy tails of which violate the sub-exponential assumption (\cref{def:sub:exp:RV}).  There appears to be no
  significant departure from the idea that the form
  $t(\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+}$ corresponds increasingly well to
  the true approximation error (\eqref{eq:true:diff}) as the dimension
  increases.
\end{remark}

@@

In addition to these scientific questions of interest, we are
interested in the performance of the first-order approximation as a
surrogate gradient for HMC in such a high-dimensional setting. Whereas we know that the surrogate-trajectory HMC transition kernel leaves its target distribution invariant regardless of the approximation quality \cite{glatt2020accept}, transitions that rely on poor gradient approximations result in small acceptance rates, more random walk behavior and high autocorrelation between samples.  Since ESS is inversely proportional to a Markov chain's asymptotic autocorrelation, larger ESS suggest a useful gradient approximation.  To
isolate the approximation's performance, we fix the Bayesian bridge global-scale
$\tau$ at 2.5$\times10^{-5}$.  We generate a Markov chain with 80,000
states, saving every tenth state and removing the first 1,000 states as
burn-in.  Despite the relatively small number of iterations, we
observe large ESS that suggest satisfactory accuracy of the
first-order approximation within the context of high-dimensional
surrogate HMC.   The ESS for the three fixed-effect regression
coefficients $\theta_1$, $\theta_2$ and $\theta_3$ are 1,053, 1,343
and 498, respectively.  The median and minimum ESS for the 1,892
random effects $\epsilon_{ij}$ are 1,514 and 1,161, respectively.

### 2023-09-14 manuscript/revision_letter.tex
more revisions

\newcommand{\bbP}{{\mathbb P}}
\newcommand{\bbE}{{\mathbb E}}
\begin{reply}
	You are correct that the truncated Cauchy (at 0) and the folded Cauchy are the same, but this distribution is not sub-exponential as we have defined the term in Definition \ref{def:sub:exp:RV}:
	\begin{quote}
\begin{definition}
	\label{def:sub:exp:RV}
	A random variable $X$ is called \textit{sub-exponential} if there
	exists some constant $K > 0$ for which its tails satisfy
	\begin{equation}\label{e:def_subexp_RV}
		\bbP(|X|\geq t) \leq 2 \exp\{-t/K\}, \quad \forall t \geq 0.
	\end{equation}
	In this case, the \textit{sub-exponential norm} of $X$ is defined by
	\begin{equation}\label{e:sub-exp_norm}
		\|X\|_{\psi_1} = \inf\big\{s > 0: \bbE \exp(|X|/s) \leq 2\big\}.
	\end{equation}
	The \textit{class of sub-exponential distributions} is denoted by
	\begin{align*}
		L_{\psi_1} = \big\{F_X(dx): \|X\|_{\psi_1} < \infty \big\}.
	\end{align*}
\end{definition}
	\end{quote}
Here, sub-exponential means the opposite of heavy-tailed.  The folded Cauchy distribution represents a distinct failure of the sub-exponential assumption.  In order to avoid further confusion, we have added the following sentence prior to Definition 1:
\begin{quote}
	To avoid confusion, note that the following definition uses the term in the same way as, e.g., \cite{vershynin:2018}, but that other definitions that mean quite the opposite (i.e., heavier than exponential tails) appear in the literature \citep{goldie1998subexponential}.
\end{quote}
Additionally, we have added the following sentence to Remark 4:
\begin{quote}
	The third plot involves folded Cauchy random variables, the heavy tails of which violate the sub-exponential assumption (Definition \ref{def:sub:exp:RV}).
\end{quote}
\end{reply}

* Equations 12-19 seems to follow straightforwardly from the section where equation 2-7 are presented. I agree that writing them explicitly is better, but I felt that they disrupt the flow of the section "Deterministic bounds on approximation error in time". Please consider emulating them to somewhere in 2-7 or following, intermediate steps in 12-19 maybe put in an appendix.

@@

@@

@@

>>>>

\newcommand{\bbP}{{\mathbb P}}
\newcommand{\bbE}{{\mathbb E}}
\begin{reply}
	You are correct that the truncated Cauchy (at 0) and the folded Cauchy are the same, but this distribution is not sub-exponential as we have defined the term in Definition \ref{def:sub:exp:RV}:
	\begin{quote}

\setcounter{equation}{39}
\begin{definition}
	\label{def:sub:exp:RV}
	A random variable $X$ is called \textit{sub-exponential} if there
	exists some constant $K > 0$ for which its tails satisfy
	\begin{equation}\label{e:def_subexp_RV}
		\bbP(|X|\geq t) \leq 2 e^{-t/K}, \quad \forall t \geq 0.
	\end{equation}
	In this case, the \textit{sub-exponential norm} of $X$ is defined by
	\begin{equation}\label{e:sub-exp_norm}
		\|X\|_{\psi_1} = \inf\big\{s > 0: \bbE e^{|X|/s} \leq 2\big\}.
	\end{equation}
	The \textit{class of sub-exponential distributions} is denoted by
	\begin{align*}
		L_{\psi_1} = \big\{F_X(dx): \|X\|_{\psi_1} < \infty \big\}.
	\end{align*}
\end{definition}
	\end{quote}
Here, sub-exponential means the opposite of heavy-tailed.  The folded Cauchy distribution represents a distinct failure of the sub-exponential assumption.  In order to avoid further confusion, we have added the following sentence prior to Definition 1:
\begin{quote}
	To avoid confusion, note that the following definition uses the term in the same way as, e.g., \cite{vershynin:2018}, but that other definitions that mean quite the opposite (i.e., heavier than exponential tails) appear in the literature \citep{goldie1998subexponential}.
\end{quote}
Additionally, we have added the following sentence to Remark 4:
\begin{quote}
	The third plot involves folded Cauchy random variables, the heavy tails of which violate the sub-exponential assumption (Definition \ref{def:sub:exp:RV}).
\end{quote}
\end{reply}

* Equations 12-19 seems to follow straightforwardly from the section where equation 2-7 are presented. I agree that writing them explicitly is better, but I felt that they disrupt the flow of the section ``Deterministic bounds on approximation error in time''. Please consider emulating them to somewhere in 2-7 or following, intermediate steps in 12-19 maybe put in an appendix.

\begin{reply}
We thank you for the suggestion, but we would really like to keep 12-19 where they are if it's alright.  We do not want to place them directly following 2-7 because we would prefer to keep the Introduction at a high level, and we do not want to put them in the Appendix because we think they communicate valuable insight into the dynamical systems view of our formal argument.  We hope you understand.
\end{reply}

@@

\begin{reply}
	Thank you for catching our inconsistent use of $\exp$.  We have updated the manuscript accordingly.  You are also correct that Vershynin provides this definition.  We have re-checked and there are no additional regularity conditions attached.   We wish to proceed with this definition, but please note that there exists a definition of sub-exponential based on CDFs that actually means heavier-tailed than exponential \citep{goldie1998subexponential}.  We point this out again to help avoid confusion.
\end{reply}

@@

\begin{reply}
	We thank you for pointing this out. We have changed the sentence to the following:
	\begin{quote}
		There appears to be no
		significant departure from the idea that the form
		$t(\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+}$ corresponds increasingly well to
		the true approximation error \eqref{eq:true:diff} as the dimension
		increases.
	\end{quote}
\end{reply}

@@

\begin{reply}
	Thank you for pointing this out.  We have added the following sentences to the paragraph:
	\begin{quote}
		Whereas we know that the surrogate-trajectory HMC transition kernel leaves its target distribution invariant regardless of the approximation quality \citep{glatt2020accept}, transitions that rely on poor gradient approximations result in small acceptance rates, more random walk behavior and high autocorrelation between samples.  Since ESS is inversely proportional to a Markov chain's asymptotic autocorrelation, larger ESS suggest a useful gradient approximation.
	\end{quote}
\end{reply}
